Ornaments as indicators of social changes in northeastern Taiwan before and after the European colonial period

### 2020-07-08 analysis/paper/paper.Rmd
edit discussion section based on R3 comment

Ornaments (Figure \@ref(fig:ornaments-image)) were found in a variety of archaeological contexts including post-holes area, burials, and middens. For the ornament categories, we follow the well-established topology based on raw materials and shapes for the region of northeastern Taiwan [@Chen2007; @NMTH2005]. This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the main habitation areas of Kiwulan, indicated by aligned post-holes with _in-situ_ posts (Figure \@ref(fig:KWL-sam-area)). Occupation floors were not identified during excavation. We choose these units because they were stratigraphically intact and undisturbed by modern construction activity, compared to excavation squares on the periphery of the site. There are 35 burials in the sampling area, one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, with the total number of ornaments in burials including 3,173. There are also 27 ornaments found in midden contexts. The obvious difference in number in burials and midden contexts indicates the property of ornaments as prestige goods. The high number in burials is due to the presence of bead strands or patterned bands of beads, which sometimes contain thousands of beads in an individual burial [@Chen2007]. The highly uneven distribution of beads between burials was interpreted as the evidence for hierarchy by @Cheng2008. However, @Hsieh2012 suggests a relatively egalitarian society from the comparative analysis of frequency and proportion of burial goods.

@@

Figure \@ref(fig:plot-altogether) shows the distribution of frequencies for subtypes in each major class. Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of the effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bells and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells have less variety, typically >2 cm long with a wide variety of human faces as a motif. Although glass beads have less variety in size, presenting as small (0.5-1 cm) or medium (1-2 cm), they have a wide variety of colors or patterns mostly made by a winding technique and with high lead content indicating possibly from China [@Cheng2008]. Although we are not certain of the specific origin of all kinds of beads, research suggest that glass beads and metal ornaments have similar production techniques and composition to those found in China, while gold-foil beads may relate to the economic activities of the Spanish in the South China Sea region [@Chen2011; @Wang2018]. There seem to be no obvious changes in the sources of glass beads or metal ornaments at different phases in the upper component of Kiwulan (1350-1850 AD). However, the glass beads from the lower component (650-1150 AD) demonstrate the composition of m-Na-Al glass and soda plant ash glass that are subtypes of the Indo-Pacific beads, frequently seen in Southeast Asia[@Carter2016; @Francis2002; @Wang2014]. A transition in the origin of glass beads from mainly Southeast Asia to multiple regions including China might indicate different trading networks. A regional network in South China Sea in the earlier period and later involved in larger scale trading network stimulated by the Europeans and the Chinese [@Hung2016; @Wang2007].

@@

# plot spatial pattern for all ornaments
plot_spatial_distribution_all_ornaments_by_period <- 
  ornaments_period_join %>% 
  group_by(period, Pit_No, geometry) %>% 
  summarize(n = sum(n)) %>% 
  ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_all_type_across_periods, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9) +
  facet_wrap(~ period) +
  labs(fill = "frequency") +
  scale_fill_viridis(direction = -1) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        legend.key.size =  unit(3, "mm"))

@@

Archaeological contexts show that ornaments are especially abundant in burial contexts serving as grave goods [@Chen2007]. This supports the interpretation of ornaments as valuable objects functioning as status indicators. Spatial patterns of ornaments in dwelling contexts show that their distribution was clustered during the pre-European and European periods. These clusters are non-random, and are most highly concentrated during the European period. This may indicate that a degree of social inequality based on the uneven distribution of ornaments was already present before European contact, and then it was reinforced and amplified during the European period. A further indicator of increased social inequality is a burial dated to the 17th century that included 60 gold-foil beads, well above the average of 2-3 pieces in the pre-European period [@Chen2007; @Cheng2008]. @Cheng2008 proposed a more hierarchical structure of Kiwulan based on the found of burial with 60 gold-foil beads that were treated as prestige goods in local community. However, @Hsieh's research on data from burial contexts suggested a more egalitarian society since burials with rich goods are elders that indicates accumulated wealth over time instead of inherited. Our results focusing on the residential area demonstrated an uneven distribution of prestige ornaments during the European presence that inclined an appearance of social inequality. Nevertheless, we still need a comprehensive exploration of other archaeological records to discuss the social structure of Kiwulan.

@@

That said, the evidence from Kiwulan may be consistent with a variety of scenarios of indigenous-colonial relations. The increasing number and concentrated spatial patterns of ornaments may also indicate a practice of cultural resistance against the European intrusion. Resistance to European economic and political demands may be inferred if ornaments were used as a display of social identity and to emphasize the local customs that had existed before European contact [cf. @Rubertone2000]. Resistance could be presented in many forms, but we have no specific evidence from Kiwulan to prefer resistance as the primary mechanism behind the distribution of ornaments over colonial influence. Another scenario is that ornaments were treated as heirlooms, such as carnelian beads and gold-foil beads, that passed from one generation to the next, accumulating at Kiwulan over time. This process would result in a natural increase in ornaments over time, unrelated to colonial influences. As noted above, this pattern is not a good fit for the ornament distribution at Kiwulan, and adds little value in explaining the shifts in spatial patterns.

@@

Micro-scale analysis of the archaeological record at the peripheries of colonial activity offers a opportunity to understand pervasive but not necessarily dominant European colonial processes at remote Indigenous groups [@Trabert2018]. Similar impacts can be also seen in modern societies today when adopting western products [@AlGhanim2017; @Zheng2015]. Kiwulan in northeastern Taiwan is an exceptional case study as an East Asian location that was relatively isolated and peripheral, and yet connected by regional and global trade networks. Kiwulan provides valuable insights into the discussion of indirect colonial influence on local societies living beyond the reach of direct European colonial occupation. The frequency and spatial distribution of personal ornaments at Kiwulan present three distinct patterns during different dominant culture interaction periods. The greater amount and diversity of ornament types during the European period reflects an increasing use in ornaments in a colonial context. Before European contact, ornaments were traded into local Indigenous societies via the regional exchange network with Chinese merchants, and viewed as prestige goods in the local Indigenous culture. After the arrival of the Europeans, the exotic and powerful image carried by those ornaments may have intensified, further signaling wealth and privileged trading connections among the inhabitants of Kiwulan. This may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation at Kiwulan, which might have resulted in an increase in social inequality. This might also indicate an act of intentional resistance to the intrusion of the Europeans by using more ornaments that are symbolic of cultural tradition, but additional evidence is required to confirm this.

By focusing on the distribution patterns in a settlement site, the Kiwulan ornaments suggest that foreign ornaments can be a proxy to detect indirect colonial influence on local Indigenous populations. Ornaments give insights into the amplification of social inequality stimulated by European colonization. It also shows the agency of Indigenous people to incorporate ornaments into their social system and use them in their daily lives to display or intensify status differences. We are still far from understanding the full variety of colonial impacts on peripheral Indigenous communities. We have introduced here the corporate/network model for understanding the dynamics of social inequality at Kiwulan, and further provenance analysis of imported ceramics and ornaments such as X-ray fluorescence analysis would provide more information to construct a clear picture of complex trade networks during this periods.

>>>>

Ornaments (Figure \@ref(fig:ornaments-image)) were found in a variety of archaeological contexts including post-holes area, burials, and middens. For the ornament categories, we follow the well-established topology based on raw materials and shapes for the region of northeastern Taiwan [@Chen2007; @NMTH2005]. This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the main habitation areas of Kiwulan, indicated by aligned post-holes with _in-situ_ posts (Figure \@ref(fig:KWL-sam-area)). Occupation floors were not identified during excavation. We choose these units because they were stratigraphically intact and undisturbed by modern construction activity, compared to excavation squares on the periphery of the site. There are 35 burials in the sampling area, one third of the total number of burials at Kiwulan. Intact ornaments are commonly found in burials used as personal adornment, indicating the property of ornaments as prestige goods. In general, there are some differences between the number of gold-foil beads, carnelian beads, and glass beads across burials. For example, over a thousand beads are found in a few burials, but none is found in some burials. This is also because of the presence of bead strands or patterned bands of beads, which sometimes contain thousands of beads in an individual burial [@Chen2007]. The uneven distribution of beads indicate some social differentiation, but a full discussion of the burial data is beyond the scope of this paper.

@@

Figure \@ref(fig:plot-altogether) shows the distribution of frequencies for subtypes in each major class. Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of the effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bells and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells have less variety, typically >2 cm long with a wide variety of human faces as a motif. Although glass beads have less variety in size, presenting as small (0.5-1 cm) or medium (1-2 cm), they have a wide variety of colors or patterns mostly made by a winding technique and with high lead content indicating possibly from China [@Cheng2008].

Although we are not certain of the specific origin of all kinds of beads, research suggest that glass beads and metal ornaments have similar production techniques and composition to those found in China, while gold-foil beads may relate to the economic activities of the Spanish in the South China Sea region [@Chen2011; @Wang2018]. There seem to be no obvious changes in the sources of glass beads or metal ornaments at different phases in the upper component of Kiwulan (1350-1850 AD). However, the glass beads from the lower component (650-1150 AD) demonstrate the composition of m-Na-Al glass and soda plant ash glass that are subtypes of the Indo-Pacific beads, frequently seen in Southeast Asia[@Carter2016; @Francis2002; @Wang2014]. A transition in the origin of glass beads from mainly Southeast Asia to multiple regions including China might indicate different trading networks. A regional network in South China Sea in the earlier period and later involved in a larger scale trading network stimulated by the Europeans and the Chinese [@Hung2016; @Wang2007].

@@

# create df for all types of ornaments by units and period
ornaments_period_join_all_type_across_periods <-
  ornaments_period_join %>% 
  group_by(period, Pit_No, geometry) %>% 
  summarize(n = sum(n))

# plot spatial pattern for all ornaments
plot_spatial_distribution_all_ornaments_by_period <- 
  ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_all_type_across_periods, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9) +
  facet_wrap(~ period) +
  labs(fill = "frequency") +
  scale_fill_viridis(direction = -1) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        legend.key.size =  unit(3, "mm"))

@@

Archaeological contexts show that ornaments are especially abundant in burial contexts serving as grave goods [@Chen2007]. This supports the interpretation of ornaments as valuable objects functioning as status indicators. Spatial patterns of ornaments in dwelling contexts show that their distribution was clustered during the pre-European and European periods. These clusters are non-random, and are most highly concentrated during the European period. This may indicate that a degree of social inequality based on the uneven distribution of ornaments was already present before European contact, and then it was reinforced and amplified during the European period. A further indicator of increased social inequality is a burial dated to the 17th century that included 60 gold-foil beads, well above the average of 2-3 pieces in the pre-European period [@Chen2007; @Cheng2008]. Based on the finding of that burial, @Cheng2008 proposed a more hierarchical structure of Kiwulan. However, @Hsieh's research on burial data suggested a more egalitarian society since a few burials with rich goods are elders that indicates accumulated wealth over time instead of inherited status. Despite the data they used for making inferences are all based on burial contexts, the inferences about Kiwulan social structure are opposite. Our results can provide an additional insight by focusing on the residential area that demonstrated an uneven distribution of ornaments during the European presence, indicating an increasing social inequality. Nevertheless, we still need to explore and compare with other archaeological records comprehensively discuss the social structure of Kiwulan.

@@

That said, the evidence from Kiwulan may be consistent with a variety of scenarios of indigenous-colonial relations. The increasing number and concentrated spatial patterns of ornaments may also indicate a practice of cultural resistance against the European intrusion. Resistance to European economic and political demands may be inferred if ornaments were used as a display of social identity and to emphasize the local customs that had existed before European contact [cf. @Rubertone2000]. We know the custom of wearing ornaments in Indigenous societies for attending ceremonial events based on the Chinese historical records in the 19th century []. Also, an ethnographic photograph clearly presents the weaving practice by a local woman, who wore a string of beads as a necklace [@Mackey1895, pp. 206-307]. Displaying ornaments in social events or daily live could be a cultural symbol, and may be reinforced during the foreign presence. This could be a form of resistance, but we still need more specific evidence from Kiwulan to prefer resistance as the primary mechanism behind the distribution of ornaments over colonial influence. Another scenario is that ornaments were treated as heirlooms, such as carnelian beads and gold-foil beads, that passed from one generation to the next, accumulating at Kiwulan over time. This process would result in a natural increase in ornaments over time, unrelated to colonial influences. As noted above, this pattern is not a good fit for the ornament distribution at Kiwulan, and adds little value in explaining the shifts in spatial patterns.

@@

Analyses of the archaeological record at the peripheries of colonial activity offers a opportunity to understand pervasive but not necessarily dominant European colonial processes at remote Indigenous groups [@Trabert2018]. Similar impacts can be also seen in modern societies today when adopting western products [@AlGhanim2017; @Zheng2015]. Kiwulan in northeastern Taiwan is an exceptional case study as an East Asian location that was relatively isolated and peripheral, and yet connected by regional and global trade networks. Kiwulan provides valuable insights into the discussion of indirect colonial influence on local societies living beyond the reach of direct European colonial occupation. The frequency and spatial distribution of personal ornaments at Kiwulan present three distinct patterns during different dominant culture interaction periods. The greater amount and diversity of ornament types during the European period reflects an increasing use in ornaments in a colonial context. Before European contact, ornaments were traded into local Indigenous societies via the regional exchange network with Chinese merchants, and viewed as prestige goods in the local Indigenous culture. After the arrival of the Europeans, the exotic and powerful image carried by those ornaments may have intensified, further signaling wealth and privileged trading connections among the inhabitants of Kiwulan. This may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation at Kiwulan, which might have resulted in an increase in social inequality. This might also indicate an act of intentional resistance to the intrusion of the Europeans by using more ornaments that are symbolic of cultural tradition, but additional evidence is required to confirm this.

By focusing on the distribution patterns in a settlement site, the Kiwulan ornaments suggest that foreign ornaments can be a proxy to detect indirect colonial influence on local Indigenous populations. Ornaments give insights into the amplification of social inequality stimulated by European colonization. It also shows the agency of Indigenous people to incorporate ornaments into their social system and use them in their lives or social occasions to display or intensify status differences. We are still far from understanding the full variety of colonial impacts on peripheral Indigenous communities. We have introduced here the corporate/network model for understanding the dynamics of social inequality at Kiwulan, and further provenance analysis of imported ceramics and ornaments such as X-ray fluorescence analysis would provide more information to construct a clear picture of complex trade networks during this periods.
